---
title: "HW6 (2025-11-19), Math 261A, Isis Martinez, ID 018233605"
output: 
  pdf_document:
    latex_engine: xelatex
    
---

**Problem 1:**

```{r, echo=FALSE, results='markup', message=FALSE, warning=FALSE}
library(tidyverse)
tornadoes <- readr::read_csv(paste0('https://raw.githubusercontent.com/',
'rfordatascience/tidytuesday/main/',
'data/2023/2023-05-16/tornados.csv')) |>
filter(yr >= 2007, !is.na(loss))
```

**Problem 1a**: A tornado's Enhanced Fujita rating is a set of wind estimates determined by estimating wind speeds and expected related damage. It is inferred from damage assessments calibrated to expected wind speeds (not measured directly by instruments). This is found by surveying tornado-related damage and comparing it to a list of Damage Indicators (DIs/28 types of structures or vegetation that could be damaged) and Degrees of Damage (DoD). Each level corresponds to an estimated range of wind speeds to assign a range of wind speeds the tornado likely produced and the rating from EF0 to EF5 follows:

Here is the EF scale as stated by the [National Weather Service website](https://www.weather.gov/oun/efscale):
**EF Rating	3 Second Gust (mph)**:

0	65-85

1	86-110

2	111-135

3	136-165

4	166-200

5	Over 200


**Problem 1b**:
We fit a simple linear regression model 
$loss_i = \beta_0 + \beta_1mag_i + \varepsilon_i$ where loss is the estimated property loss in dollars from a tornado, and mag is the tornado's EF rating.

```{r, echo=TRUE, results='markup', message=FALSE, warning=FALSE}
lm_fit <- lm(data=tornadoes, loss ~ mag)
summary(lm_fit)
slope <- coef(lm_fit)["mag"]

```
The slope parameter $\beta_1$ is the estimated change in change in USD property loss for every 1-unit increase in the tornado’s EF rating. As we see from the summary above, the slope of this model is \$ `r round(slope, 3)`.

**Problem 1c**: We fit a  log linear regression model $log(loss_i) = \beta_0 + \beta_1mag_i + \varepsilon_i$:

```{r, echo=TRUE, results='markup', message=FALSE, warning=FALSE}

log_fit <- lm(data=tornadoes, log(loss) ~ mag)
summary(log_fit)
log_slope <- coef(log_fit)["mag"]

```

The estimated slope parameter is $b_1 \approx$ `r round(log_slope, 3)`, which means every 1 unit change in a tornado's EF score corresponds to an expected average difference of log(loss) in damage in USD. 

In other words, the expected property loss in US dollars changes by approximately $100 \times (e^{b_1} - 1) =$ `r round(100 * (exp(log_slope) - 1), 1)`% since $e^{x_i} \approx 1+x_i$.

The slope from part (b) represented a change in dollars of damage, while this logarithmic model 's slope corresponds to approximate proportional differences represents.

The log transformation normalizes the loss distributions and makes the relationship more linear, so the log-linear model is preferred.



**Problem 2**: The below code simulates training and testing data for two variables X and Y from the model $Y= 2X + \varepsilon$, where $\varepsilon$ is not constant. In particular, the standard deviation of the errors $\varepsilon$ is dependent on $X$, and is equal to $1 + x^2/2$. The code below also provides examples of how to fit the linear regression model $Y\sim X$ using ordinary least squares and weighted least squares.


**Problem 2a**: First, show that if $E(Y) = X\beta$, the weighted least squares estimator is unbiased.

We know that the weighted least squares is $\hat{\beta}_{WLS}=(X^TWX)^{-1}X^TWy$, where W is a W is a symmetric positive-definite weight matrix. We know it's unbiased if $E(\hat{\beta}_{OLS})=\beta$. Therefore, we compute the following:

$E[\hat{\beta}_{WLS}]=E[(X^TWX)^{-1}X^TWY] = (X^TWX)^{-1}X^TWE[Y]=(X^TWX)^{-1}X^TWX\beta]$

Since $(X^TWX)^{-1}$ is the inverse of $X^TWX$ with W being symmetric, we therefore can conclude $E[\hat{\beta}_{WLS}]=(X^TWX)^{-1}X^TWX\beta]=\beta$. So, we know that $\hat{\beta}_{WLS}$ is unbiased.


**Problem 2b**
Use the below code as a starting point for a simulation study that shows that the weighted least squares estimator $\hat{\beta}^{WLS}_1$ is more efficient than the OLS estimator (meaning that $SE(\hat{\beta}^{WLS}_1)$ is less than $SE(\hat{\beta}^{OLS}_1)$).

```{r, echo=TRUE, results='markup', message=FALSE, warning=FALSE}
set.seed(123)
n <- 50

sims <-500
# containers
b1_ols  <- numeric(sims)
b1_wls  <- numeric(sims)
se1_ols <- numeric(sims)
se1_wls <- numeric(sims)

for (s in 1:sims) {
  #train and test
  X_train <- runif(n, min = -5, max = 5)
  X_test <- runif(n, min = -5, max = 5)

  Y_train <- 2 * X_train + rnorm(n, 0, sd = 1 + X_train ^ 2 / 2)
  Y_test <- 2 * X_train + rnorm(n, 0, sd = 1 + X_train ^ 2 / 2)
  
  # fit models
  fit_ols = lm(Y_train ~ X_train)
  fit_wls = lm(Y_train ~ X_train, weights = 1/(1 + 0.5 * X_train^2) ^ 2)
  
  b1_ols[s]  <- coef(fit_ols)[2]
  b1_wls[s]  <- coef(fit_wls)[2]
  se1_ols[s] <- summary(fit_ols)$coef[2, 2]
  se1_wls[s] <- summary(fit_wls)$coef[2, 2]
}

#sampling SDs of slope estimates
samp_sd_ols <- sd(b1_ols)
samp_sd_wls <- sd(b1_wls)

# avg model-reported SEs for the slope
avg_se_ols <- mean(se1_ols)
avg_se_wls <- mean(se1_wls)

#relative efficiency (if number >1, WLS more efficient)
rel_eff <- (samp_sd_ols^2) / (samp_sd_wls^2)

cat(sprintf("Observed SD of slope (OLS): %.4f\n", samp_sd_ols))
cat(sprintf("Observed SD of slope (WLS): %.4f\n", samp_sd_wls))
cat(sprintf("Relative efficiency Var(OLS)/Var(WLS): %.2f\n\n", rel_eff))
cat(sprintf("Average reported SE (OLS): %.4f\n", avg_se_ols))
cat(sprintf("Average reported SE (WLS): %.4f\n", avg_se_wls))

```
The observed standard deviation of the OLS estimates across all simulations is 0.4356. It measures how much the OLS slope estimate varies between samples. 

The observed standard deviation of the OLS estimates across all simulations is 0.2699, which is less than 0.4356. This means the WLS estimates with less variability across repeated samples. Therefore, we see that the WLS slope estimates are less noisy/more stable than the OLS estimates, which is a sign of higher efficiency.

We also measured the observed relative efficiency Var(OLS)/Var(WLS), which was 2.60. If $0> efficiency < 1$, then that would indicate $SE(\hat{\beta}^{WLS}_1) >SE(\hat{\beta}^{OLS}_1)$. However, since this value is greater than 1, we know that $SE(\hat{\beta}^{WLS}_1)$ is less than $SE(\hat{\beta}^{OLS}_1)$, and therefore the WLS estimator is more efficient.

**Problem 2c**: Use the below code as a starting point for a simulation study that compares the mean squared prediction error of the OLS and WLS estimators. Which estimator yields the better test MSE, on average?

```{r, echo=TRUE, results='markup', message=FALSE, warning=FALSE}
set.seed(123)
n <- 50

sims <-500
# containers
mse_ols <- numeric(sims)
mse_wls <- numeric(sims)

for (s in 1:sims) {
  #train and test
  X_train <- runif(n, min = -5, max = 5)
  X_test <- runif(n, min = -5, max = 5)

  Y_train <- 2 * X_train + rnorm(n, 0, sd = 1 + X_train ^ 2 / 2)
  Y_test <- 2 * X_train + rnorm(n, 0, sd = 1 + X_train ^ 2 / 2)
  
  # fit models
  fit_ols = lm(Y_train ~ X_train)
  fit_wls = lm(Y_train ~ X_train, weights = 1/(1 + 0.5 * X_train^2) ^ 2)
  
  # predictions on the test X's
  yhat_ols <- predict(fit_ols, newdata = data.frame(X_train = X_test))
  yhat_wls <- predict(fit_wls, newdata = data.frame(X_train = X_test))

  # test MSEs
  mse_ols[s] <- mean((Y_test - yhat_ols)^2)
  mse_wls[s] <- mean((Y_test - yhat_wls)^2)
}

avg_mse_ols <- mean(mse_ols)
avg_mse_wls <- mean(mse_wls)
sd_mse_ols  <- sd(mse_ols)
sd_mse_wls  <- sd(mse_wls)

#relative efficiency (if number >1, WLS more efficient)
rel_eff <- (samp_sd_ols^2) / (samp_sd_wls^2)

cat(sprintf("Average test MSE (OLS): %.4f  (SD across sims: %.4f)\n", avg_mse_ols, sd_mse_ols))
cat(sprintf("Average test MSE (WLS): %.4f  (SD across sims: %.4f)\n", avg_mse_wls, sd_mse_wls))
cat(sprintf("Average difference (OLS - WLS): %.4f\n", mean(mse_ols - mse_wls)))


```
We see that the WLS has a slightly smaller average test MSE, and the WLS test errors vary a bit less from run to run than the OLS test errors.

Both methods perform similarly in terms of prediction error, since the mean test MSEs are very close (108.2 vs. 107.3). That's expected, because both models are correctly specified since they estimate the same true mean function $E[Y\mid X]=2X$. WLS performs slightly better on average with the difference being $\approx 0.94$. It is small but consistent with the what we would expect that when heteroskedasticity is present and weights correctly reflect the inverse of the error variance, WLS produces slightly more precise coefficient estimates. Meaning overall, on average, WLS achieves slightly lower and more consistent test MSE, reflecting its ability to account for non-constant error variance.


**Problem 3a**:
29 independent analysis teams with total 61 analysts explored the same research question with the same dataset: are soccer referees more likely to give red cards to dark-skin-toned players than to light-skin-toned players?

The dataset and question were fixed, but the analytic path was open since each team could make its own decisions about data cleaning, the covariates to include, model specification, exclusions, etc. 

The authors found effect estimates ranged from 0.89 to 2.93. 20 out of the 29 researchers ($\approx 69%$) of teams found a statistically significant positive effect (dark-skin-toned players more likely penalized), while the other 9 ($\approx 31%$) did not. The analyses used 21 unique combinations of covariates across the 29 teams showing differences in analytic approaches. 

The study shows that analytic choices can produce differing results, even with identical data. The authors recommend greater transparency, reporting of analytic decisions, and multi-analyst or multiple approaches to show how robust findings actually are.

**Problem 3b**. We look at the research question given: Are soccer referees more likely to give red cards to players with darker skin tones, after accounting for game context and player characteristics?

The dataset records the **number of red cards** per player across matches.

Here are the details of the proposed model:

* **Response variable**:  Y = number_of_red_cards. This variable can be aggregated per player or per match. Red cards are non-negative integers, a Poisson or log-linear model is could work.

* **Predictor variables**: 

  - skin_tone (categorical: dark vs. light) - primary variable of interest
  - position (categorical: defender, midfielder, forward) - controls for differing defensive risk
  - fouls_committed (numeric) - proxies player aggression
  - minutes_played (numeric) - exposure variable; included as offset
  - team_strength or opponent_strength (numeric or categorical) - contextual control
  - referee_id or league (random or fixed effects) - accounts for variation among referees/leagues
    
**Proposed model**: A mixed logistic-linear model:

$$\log(E[Y_i]) = \beta_0 + \beta_1 (SkinTone_i) + \beta_2 (position_i) + \beta_3 (FoulsCommitted_i) + \beta_4 (TeamStrength_i) + \log(MinutesPlayed_i)$$
where the offset term adjusts for different exposure times (minutes played).


Reasoning:

* The log-linear model could be appropriate because the number of red cards discrete data that typically follows a a Poisson-like distribution.
* Including minutes played as an control helps make sure players with more playing time are appropriately accounted for.
* Covariates like position, fouls committed, and team strength control for performance and context differences unrelated to referee bias.
* The coefficient for skin_tone quantifies whether darker-skinned players receive unbalanced number of red cards.

Example code for implementation after loading libraries and dataset:

Fit log-linear (Poisson) regression
model <- glm(
  red_cards ~ skin_tone + position + fouls_committed + team_strength + offset(log(minutes_played)),
  data = data,
  family = poisson(link = "log"))
summary(model)

Interpretation: A positive coefficient for skin_tone (dark) would suggest that dark-skin-toned players are more likely to receive red cards, controlling for other factors if it is deemed to be significant. The exponentiated coefficients (using exp(coef(model))) provide rate ratios, which is the proportional change in expected red-card count per predictor unit.